---
title: "Mixed effects models for movement"
output: 
    github_document: default
---

## Set up
```{r}
library(nlme)
library(lme4)
library(data.table)
library(ggplot2)
library(RColorBrewer)
dat <- fread('data/data_speed.csv')
```
## Simulated data
```{r}
params <- data.frame(a = 5, b = 2, w = 5)
sim <- data.table(temperature = sample(10:30, 125, replace = TRUE),
                  species = as.factor(sample(c('a', 'b', 'c', 'd', 'e', 'f', 'g', 'h'), 125, replace = TRUE)))
sim[, mass := pmax(20, as.numeric(species) * 100 + rnorm(.N, 0, 50))]
sim[, log.mass.mg := log(mass)]
sim[, z := as.numeric(species)*2 + 10]
sim[, log.speed := params$a + params$b * log.mass.mg - (temperature -z)^2/params$w + rnorm(.N, 0, 2)]
ranefs = sim[, .(z = unique(z)), by = species]

ggplot(sim, aes(temperature, log.speed, color = species)) +
    geom_point()
```

### A first model on simulated data
```{r}
metabfunc <- function(lmass, temp, a, b, z, w){
    lspeed <- a + b * lmass - (temp - z)^2/w
}
modsim1 <- nlme(log.speed ~ metabfunc(log.mass.mg, temperature, a, b, z, w), 
             random = z ~ 1|species,
             fixed = a + b + w ~ 1,
             start = list(fixed = c(a = 4.8, b = 1.8, w = 4.8)),
             data = sim)

summary(modsim1)
```

### Compare estimates to true values
```{r}
fixef_est <- summary(modsim1)$tTable
ranef_est <- ranef(modsim1)
ranef_est$species <- rownames(ranef_est)
colnames(ranef_est) <- c('z_est', 'species')

print(data.frame(True = t(params), Est = fixef_est[, 'Value'], StdErr = fixef_est[, 'Std.Error']))

print(merge(ranefs, ranef_est))
```

## Real data
### Plot
```{r, echo = FALSE}
ggplot(dat, aes(temperature, log.speed, color = species)) +
    geom_point()

```

### Original model
```{r}
modorig <- lmer(log.speed ~ log.mass.mg + temperature + I(temperature^2) + (1|chamber:year),
                data = dat)

summary(modorig)
```
### Initial nlme model
Has random intercepts for chamber and year, like lmer model
```{r}
metabfunc1 <- function(lmass, temp, a, b, z, w, y){
    lspeed <- a + b * lmass - (temp - 20 + z)^2/w + y
}


mod0 <- nlme(log.speed ~ metabfunc1(log.mass.mg, temperature, a, b, z, w, y), 
             fixed = a + b + w + z ~ 1,
             random = y ~ 1|chamber/year,
             start = list(fixed = c(a =  -5.6, b = 0.09, w = 5, z= 0)),
             data = dat)
summary(mod0)
ranef(mod0)
```

#### Plot vs. observed
```{r, echo = FALSE}
dat[, pred_mod0 := predict(mod0)]
ggplot(dat, aes(log.speed, pred_mod0, color = species)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1)

```


#### Plot predictions
```{r, echo = FALSE}
ggplot(dat, aes(temperature, pred_mod0, color = species)) +
    geom_point()
```

### Model with random optimal temperatures
```{r}
metabfunc2 <- function(lmass, temp, a, b, z0, z, w){
    lspeed <- a + b * lmass - (temp - z0 - z)^2/w
}


mod1 <- nlme(log.speed ~ metabfunc2(log.mass.mg, temperature, a, b, z0, z, w), 
             fixed = a + b + w + z0 ~ 1,
             random = z ~ 1|species,
             start = list(fixed = c(a =  -3.2, b = 0.09, w = 100, z0= 20)),
             data = dat)
summary(mod1)
print(ranef(mod1))
```

#### Plot vs. observed
```{r, echo = FALSE}
dat[, pred_mod1 := predict(mod1)]

ggplot(dat, aes(log.speed, pred_mod1, color = species)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1)

```

#### Plot predictions
Each species' curve is shown with its T_opt plotted as a triangle at the bottom. The cross-species average is plotted in black.
```{r, echo = FALSE}
preds <- as.data.table(expand.grid(temperature = seq(5, 35, by = 0.1), 
                                   species = unique(dat$species), 
                                   chamber = unique(dat$chamber), year = unique(dat$year),
                                   mass_type = c('byspecies', 'average'),
                                   stringsAsFactors = FALSE)) # data table to predict to

# add the average mass per species to the new data table for predictions
mass <- dat[, .(log.mass.mg = mean(log.mass.mg)), by = species]
preds <- merge(preds, mass)
preds[mass_type == 'average', log.mass.mg := mean(mass$log.mass.mg)] # for half the rows, use the average mass across all species (for the fixed effect predictions)

# predict from the random-Topt model
preds[, pred_mod1 := predict(mod1, newdata = preds)]
preds[, pred_mod1_fixef := predict(mod1, newdata = preds, level = 0)] # w/out Topt by species

ranefs <- ranef(mod1) # extract the random effects to get each species' Topt
ranefs$species = rownames(ranefs)
ranefs <- as.data.table(ranefs)

cols <- brewer.pal(8, 'Dark2')
sp <- unique(preds$species)

## make the plot
plot(0,0, xlim=c(5,35), ylim = c(-5,-2), xlab = 'Temperature', ylab = 'log speed')
for(i in 1:length(sp)){ # plot a curve for each species
    minT <- dat[species == sp[i], min(temperature)]
    maxT <- dat[species == sp[i], max(temperature)]
    dat[species == sp[i], points(temperature, log.speed, col = cols[i])]
    ave <- preds[species == sp[i] & 
                     temperature >= minT & temperature <= maxT &
                     !is.na(pred_mod1) &
                     mass_type == 'byspecies', .(pred_mod1 = mean(pred_mod1)), by = temperature] # average the prediction across years/chambers (all the same anyway) for this species at its species-specific mass
    ave[, lines(temperature, pred_mod1, col = cols[i])] # plot the curve for this species at its species-specific mass
    
    
    ranefs[species == sp[i], points(fixef(mod1)[4] + z, -5, col = cols[i], pch = 2)] # plot a triangle at the Topt for this species
}
ave2 <- preds[!is.na(pred_mod1_fixef) & mass_type == 'average', .(pred_mod1_fixef = mean(pred_mod1_fixef)), by = temperature] # average the fixed-effect prediction across species (all the same anyway) with a cross-species average mass
ave2[, lines(temperature, pred_mod1_fixef, col = 'black', lwd = 2)] # plot the curve for the average species
points(fixef(mod1)['z0'], -5, col = 'black', pch = 2) # plot triangle at cross-species Topt
legend('topleft', legend = sp, pch = 1, col = cols, cex = 0.5)
```

