---
title: "Mixed effects models for movement"
output: 
    github_document: default
---

## Set up
```{r}
library(nlme)
library(lme4)
library(data.table)
library(ggplot2)
library(RColorBrewer)
dat <- fread('data/data_speed.csv')
```
## Simulated data
```{r}
params <- data.frame(a = 5, b = 2, w = 5)
sim <- data.table(temperature = sample(10:30, 125, replace = TRUE),
                  species = as.factor(sample(c('a', 'b', 'c', 'd', 'e', 'f', 'g', 'h'), 125, replace = TRUE)))
sim[, mass := pmax(20, as.numeric(species) * 100 + rnorm(.N, 0, 50))]
sim[, log.mass.mg := log(mass)]
sim[, z := as.numeric(species)*2 + 10]
sim[, log.speed := params$a + params$b * log.mass.mg - (temperature -z)^2/params$w + rnorm(.N, 0, 2)]
ranefs = sim[, .(z = unique(z)), by = species]

ggplot(sim, aes(temperature, log.speed, color = species)) +
    geom_point()
```

### A first model on simulated data
```{r}
metabfunc <- function(lmass, temp, a, b, z, w){
    lspeed <- a + b * lmass - (temp - z)^2/w
}
modsim1 <- nlme(log.speed ~ metabfunc(log.mass.mg, temperature, a, b, z, w), 
             random = z ~ 1|species,
             fixed = a + b + w ~ 1,
             start = list(fixed = c(a = 4.8, b = 1.8, w = 4.8)),
             data = sim)

summary(modsim1)
```

### Compare estimates to true values
```{r}
fixef_est <- summary(modsim1)$tTable
ranef_est <- ranef(modsim1)
ranef_est$species <- rownames(ranef_est)
colnames(ranef_est) <- c('z_est', 'species')

print(data.frame(True = t(params), Est = fixef_est[, 'Value'], StdErr = fixef_est[, 'Std.Error']))

print(merge(ranefs, ranef_est))
```

## Real data
### Plot
```{r}
ggplot(dat, aes(temperature, log.speed, color = species)) +
    geom_point()

```

### Original model
```{r}
modorig <- lmer(log.speed ~ log.mass.mg + temperature + I(temperature^2) + (1|chamber:year),
                data = dat)

summary(modorig)
```
### Initial nlme model
Has random intercepts for chamber and year, like lmer model
```{r}
metabfunc1 <- function(lmass, temp, a, b, z, w, y){
    lspeed <- a + b * lmass - (temp - 20 + z)^2/w + y
}


mod0 <- nlme(log.speed ~ metabfunc1(log.mass.mg, temperature, a, b, z, w, y), 
             fixed = a + b + w + z ~ 1,
             random = y ~ 1|chamber/year,
             start = list(fixed = c(a =  -5.6, b = 0.09, w = 5, z= 0)),
             data = dat)
summary(mod0)
ranef(mod0)
```

#### Plot vs. observed
```{r}
dat[, pred_mod0 := predict(mod0)]
ggplot(dat, aes(log.speed, pred_mod0, color = species)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1)

```


#### Plot predictions
```{r}
ggplot(dat, aes(temperature, pred_mod0, color = species)) +
    geom_point()
```

### Model with random optimal temperatures
```{r}
metabfunc2 <- function(lmass, temp, a, b, z0, z, w){
    lspeed <- a + b * lmass - (temp - z0 - z)^2/w
}


mod1 <- nlme(log.speed ~ metabfunc2(log.mass.mg, temperature, a, b, z0, z, w), 
             fixed = a + b + w + z0 ~ 1,
             random = z ~ 1|species,
             start = list(fixed = c(a =  -3.2, b = 0.09, w = 100, z0= 20)),
             data = dat)
summary(mod1)
print(ranef(mod1))
```

#### Plot vs. observed
```{r}
dat[, pred_mod1 := predict(mod1)]

ggplot(dat, aes(log.speed, pred_mod1, color = species)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1)

```

#### Plot predictions
```{r}
preds <- as.data.table(expand.grid(temperature = seq(5, 35, by = 0.1), 
                                   species = unique(dat$species), 
                                   chamber = unique(dat$chamber), year = unique(dat$year),
                                   stringsAsFactors = FALSE))
mass <- dat[, .(log.mass.mg = mean(log.mass.mg)), by = species]
preds <- merge(preds, mass)

preds[, pred_mod1 := predict(mod1, newdata = preds)]

ranefs <- ranef(mod1)
ranefs$species = rownames(ranefs)
ranefs <- as.data.table(ranefs)

cols <- brewer.pal(8, 'Dark2')
sp <- unique(preds$species)

plot(0,0, xlim=c(5,35), ylim = c(-5,-2), xlab = 'Temperature', ylab = 'log speed')
for(i in 1:length(sp)){
    minT <- dat[species == sp[i], min(temperature)]
    maxT <- dat[species == sp[i], max(temperature)]
    dat[species == sp[i], points(temperature, log.speed, col = cols[i])]
    ave <- preds[species == sp[i] & 
                     temperature >= minT & temperature <= maxT &
                     !is.na(pred_mod1), .(pred_mod1 = mean(pred_mod1)), by = temperature]
  ave[, lines(temperature, pred_mod1, col = cols[i])]
  ranefs[species == sp[i], points(fixef(mod1)[4] + z, -5, col = cols[i], pch = 2)]
}
legend('topleft', legend = sp, pch = 1, col = cols, cex = 0.5)
```

### Model with random optimal temperatures and chamber and year effects
Not working well. Assigns most of the random effects to year differences rather than species differences. Would be better to have separate random effects for the optimal temperature (by species) and for an intercept (by chamber/year).
```{r}
metabfunc3 <- function(lmass, temp, a, b, z0, z, w){
    lspeed <- a + b * lmass - (temp - z0 + z)^2/w
}


mod2 <- nlme(log.speed ~ metabfunc3(log.mass.mg, temperature, a, b, z0, z, w), 
             fixed = a + b + w + z0 ~ 1,
             random = z ~1|species/chamber/year,
             start = list(fixed = c(a =  -3.2, b = 0.09, w = 100, z0= 20)),
             data = dat)
summary(mod2)
print(ranef(mod2))
```
#### Plot predictions
```{r}
preds <- as.data.table(expand.grid(temperature = seq(5, 35, by = 0.1), 
                                   species = unique(dat$species), 
                                   chamber = unique(dat$chamber), year = unique(dat$year),
                                   stringsAsFactors = FALSE))
mass <- dat[, .(log.mass.mg = mean(log.mass.mg)), by = species]
preds <- merge(preds, mass)

preds[, pred_mod2 := predict(mod2, newdata = preds)]

cols <- brewer.pal(8, 'Dark2')
sp <- unique(preds$species)

plot(0,0, xlim=c(5,35), ylim = c(-5,-2), xlab = 'Temperature', ylab = 'log speed')
for(i in 1:length(sp)){
    minT <- dat[species == sp[i], min(temperature)]
    maxT <- dat[species == sp[i], max(temperature)]
    dat[species == sp[i], points(temperature, log.speed, col = cols[i])]
    ave <- preds[species == sp[i] & 
                     temperature >= minT & temperature <= maxT &
                     !is.na(pred_mod2), .(pred_mod2 = mean(pred_mod2)), by = temperature]
  ave[, lines(temperature, pred_mod2, col = cols[i])]
  
}
```